---
title: "PCA, CCA and SiMLR: simple examples from simulated data"
author: "Brian B. Avants"
date: "2/5/2025"
output: html_document
---

PCA

	•	Ellipse: Represents the data spread based on its covariance structure.
	•	Principal Component Vectors:
	•	Solid black arrow for PC1 (explains the most variance).
	•	Dashed black arrow for PC2 (orthogonal to PC1).

This gives an intuitive visualization of PCA in action.

```{r pca}
# Load necessary libraries
library(ggplot2)
library(ggfortify)
library(ellipse)
library(ANTsR)
library(subtyper)
library( simlr.nhanes )

# Generate sample data
set.seed(42)
data <- matrix(rnorm(200), ncol = 2)
data <- data %*% matrix(c(1, 0.4, 0.6, 1), ncol = 2)  # Introduce correlation
df <- as.data.frame(data)
colnames(df) <- c("Feature1", "Feature2")

# Perform PCA
pca_result <- prcomp(df, center = TRUE, scale. = TRUE)

# Get mean of the data
center_x <- mean(df$Feature1)
center_y <- mean(df$Feature2)

# Get principal component vectors (scaled for visualization)
pc1 <- pca_result$rotation[,1] * sqrt(pca_result$sdev[1]) * 2
pc2 <- pca_result$rotation[,2] * sqrt(pca_result$sdev[2]) * 2

# Generate ellipse to fit the data
ellipse_data <- as.data.frame(ellipse(cov(df), centre = colMeans(df)))
colnames(ellipse_data) <- c("Feature1", "Feature2")  # Rename for ggplot compatibility

# Plot PCA results with ellipse and principal component vectors
ggplot(df, aes(x = Feature1, y = Feature2)) +
  geom_point(color = 'blue', alpha = 0.6) +  # Scatter plot of data
  geom_path(data = ellipse_data, aes(x = Feature1, y = Feature2), color = 'red', linewidth = 1) +  # Elliptical fit
  annotate("segment", x = center_x, y = center_y, 
           xend = center_x + pc1[1], yend = center_y + pc1[2], 
           arrow = arrow(length = unit(0.2, "cm")), color = "black", linewidth = 1.2) +  # PC1
  annotate("segment", x = center_x, y = center_y, 
           xend = center_x + pc2[1], yend = center_y + pc2[2], 
           arrow = arrow(length = unit(0.2, "cm")), color = "black", linewidth = 1.2, linetype = "dashed") +  # PC2
  ggtitle("PCA with Elliptical Fit and Principal Component Axes") +
  theme_minimal()
```


First plot the original data (X1, X2 vs. Y1, Y2) and then display the transformed low-dimensional projections (U1 vs. V1 from CCA).

```{r cca,fig.width=10,fig.height=4}

# Load necessary libraries
library(ggplot2)
library(gridExtra)
library(ellipse)

# Set seed for reproducibility
set.seed(42)

# Generate X (two independent variables)
X <- matrix(rnorm(200), ncol = 2)

# Generate Y with weaker dependence on X
Y <- X %*% matrix(c(0.3, 0.1, 0.1, 0.3), ncol = 2) + matrix(rnorm(200, sd = 0.8), ncol = 2)

# Convert to data frames
df_X <- as.data.frame(X)
df_Y <- as.data.frame(Y)
colnames(df_X) <- c("X1", "X2")
colnames(df_Y) <- c("Y1", "Y2")

# Perform Canonical Correlation Analysis (CCA)
cca_result <- cancor(df_X, df_Y)

# Compute the first canonical variates
U1 <- as.matrix(df_X) %*% cca_result$xcoef[,1]  # First canonical variate for X
V1 <- as.matrix(df_Y) %*% cca_result$ycoef[,1]  # First canonical variate for Y

# Create a dataframe for plotting
cca_data <- data.frame(U1 = U1, V1 = V1)

# Get centers
center_x <- mean(cca_data$U1)
center_y <- mean(cca_data$V1)

# Compute canonical correlation axes
cca_axis1_x <- cca_result$xcoef[,1] * 2
cca_axis1_y <- cca_result$ycoef[,1] * 2

# Generate ellipse for canonical variates
ellipse_data <- as.data.frame(ellipse(cov(cca_data), centre = colMeans(cca_data)))
colnames(ellipse_data) <- c("U1", "V1")

# Original Data Plot
p1 <- ggplot() +
  geom_point(data = df_X, aes(x = X1, y = X2), color = 'blue', alpha = 0.6) +
  geom_point(data = df_Y, aes(x = Y1, y = Y2), color = 'red', alpha = 0.6) +
  ggtitle("Original Data (X and Y)") +
  theme_minimal()

# CCA Projection Plot
p2 <- ggplot(cca_data, aes(x = U1, y = V1)) +
  geom_point(color = 'blue', alpha = 0.6) +  # Scatter plot of canonical variates
  geom_path(data = ellipse_data, aes(x = U1, y = V1), color = 'red', linewidth = 1) +  # Elliptical fit
  annotate("segment", x = center_x, y = center_y, 
           xend = center_x + cca_axis1_x[1], yend = center_y + cca_axis1_y[1], 
           arrow = arrow(length = unit(0.2, "cm")), color = "black", linewidth = 1.2) +  # First canonical axis
  annotate("segment", x = center_x, y = center_y, 
           xend = center_x + cca_axis1_x[2], yend = center_y + cca_axis1_y[2], 
           arrow = arrow(length = unit(0.2, "cm")), color = "black", linewidth = 1.2, linetype = "dashed") +  # Second canonical axis
  ggtitle("CCA Projection (U1 vs V1) - Moderate Correlation") +
  theme_minimal()

# Arrange plots side by side
grid.arrange(p1, p2, ncol = 2)

```



now three views


```{r simlrsetup,fig.width=12,fig.height=6}

# Load necessary libraries
library(ggplot2)
library(gridExtra)
library(geigen)

# Set seed for reproducibility
set.seed(42)

nzsd=0.5
mync=c(2,225,350,1070)
mynz=c(0.1,0.2,0.3)
# Generate X (three independent variables)
latent <- matrix(rnorm(300, sd=0.25 ), ncol = mync[1])

X = matrix_from_latent( latent, mync[2], mynz[1] )
Y = matrix_from_latent( latent, mync[3], mynz[2] )
Z = matrix_from_latent( latent, mync[4], mynz[3] )


# Convert to data frames
df_X <- as.data.frame(X)
df_Y <- as.data.frame(Y)
df_Z <- as.data.frame(Z)
colnames(df_X) <- paste0("X",1:ncol(X))
colnames(df_Y) <-  paste0("Y",1:ncol(Y))
colnames(df_Z) <-  paste0("Z",1:ncol(Z))
```


```{r simlr0,fig.width=12,fig.height=6}
# --- GCCA Computation ---
data_list <- list(as.matrix(df_X), as.matrix(df_Y), as.matrix(df_Z))

# Compute covariance matrices
cov_matrices <- lapply(data_list, cov)
inv_cov_matrices <- lapply(cov_matrices, function(C) solve(C + diag(1e-4, ncol(C))))  # Regularization


regs <- regularizeSimlr(data_list,fraction=0.15,sigma=rep(1.0,3))
initu=initializeSimlr( data_list, k=20 )
if ( ! exists( "result2" ) ) {
    result2 <- simlr(data_list,regs, iterations=100, 
        sparsenessQuantiles=rep(0.5,length(data_list)),
        positivities=rep("positive",length(data_list)), 
        energyType='regression',
        mixAlg='ica',
        initialUMatrix=initu, verbose=T)
    }
pred2 <- predictSimlr(data_list, result2)
U_X <- data_list[[1]] %*% (result2$v[[1]])
U_Y <- data_list[[2]] %*% (result2$v[[2]])
U_Z <- data_list[[3]] %*% (result2$v[[3]])
   
# Create a dataframe for visualization
gcca_data <- data.frame(U1 = U_X[,1], U2 = U_X[,2], 
                        V1 = U_Y[,1], V2 = U_Y[,2], 
                        W1 = U_Z[,1], W2 = U_Z[,2])

```

```{r simlr1,fig.width=12,fig.height=6}
# X vs Y
p_XY <- ggplot(gcca_data, aes(x = U1, y = V1)) +
  geom_point(color = 'purple', alpha = 0.6) +
  ggtitle("SiMLR Projection: X vs Y") +
geom_smooth(method = "lm", color = "blue", linetype = "dashed", se = FALSE) +    theme_minimal()

# X vs Z
p_XZ <- ggplot(gcca_data, aes(x = U1, y = W1)) +
  geom_point(color = 'orange', alpha = 0.6) +
  ggtitle("SiMLR Projection: X vs Z") +
geom_smooth(method = "lm", color = "blue", linetype = "dashed", se = FALSE) +    theme_minimal()

# Y vs Z
p_YZ <- ggplot(gcca_data, aes(x = V1, y = W1)) +
  geom_point(color = 'green', alpha = 0.6) +
  ggtitle("SiMLR Projection: Y vs Z") +
geom_smooth(method = "lm", color = "blue", linetype = "dashed", se = FALSE) +  theme_minimal()

# --- Original Data Plots ---
# X Data
p_X <- ggplot(df_X, aes(x = X1, y = X2)) +
  geom_point(color = 'blue', alpha = 0.6) +
  ggtitle("Original X Data") +
  theme_minimal()

# Y Data
p_Y <- ggplot(df_Y, aes(x = Y1, y = Y2)) +
  geom_point(color = 'red', alpha = 0.6) +
  ggtitle("Original Y Data") +
  theme_minimal()

# Z Data
p_Z <- ggplot(df_Z, aes(x = Z1, y = Z2)) +
  geom_point(color = 'green', alpha = 0.6) +
  ggtitle("Original Z Data") +  theme_minimal()

# Arrange the plots: GCCA and Original Data
grid.arrange(p_X, p_Y, p_Z, p_XY, p_XZ, p_YZ, ncol = 3)

```

