---
title: "simlr: nhanes application explained"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{simlr: nhanes application explained}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---



```{r setup}
# Load necessary libraries
library(ggplot2)
library(ggfortify)
library(ellipse)
library(ANTsR)
library(subtyper)
library( nhanesA )
```


## NHANES example


The five publicly available data categories are: - Demographics (DEMO) - Dietary (DIET) - Examination (EXAM) - Laboratory (LAB) - Questionnaire (Q). The abbreviated forms in parentheses may be substituted for the long form in nhanesA commands.

INDFMPIR: Poverty Income Ratio (0 = no income, higher values = higher income relative to poverty threshold).
	•	INDHHIN2: Annual Household Income (categorized ranges).
	•	DMDEDUC2: Education Level (for adults 20+ years).
	•	OCD390G: Employment Status (in occupation questionnaire).

```{r nhanes}


# Demographic data (e.g., 2011-2012 cycle)
demo_data <- nhanes('DEMO_G')
cog_data <- nhanes('CFQ_G')
depression_data <- nhanes('DPQ_G')
sleep_data <- nhanes('SLQ_G')
bp = nhanes( "BPX_G")
bm = nhanes( "BMX_G")
env_data <- nhanes('PBCD_G')  # Example for phthalates and blood metals

# Demographic data (e.g., 2013-2014 cycle)
demo_data13 <- nhanes('DEMO_H')
cog_data13<- nhanes('CFQ_H')
depression_data13 <- nhanes('DPQ_H')
sleep_data13 <- nhanes('SLQ_H')
bp13 = nhanes( "BPX_H")
bm13 = nhanes( "BMX_H")
env_data13 <- nhanes('PBCD_H')  # Example for phthalates and blood metals


phdata <- nhanes('PHTHTE_G')  # Example for phthalates and blood metals
phdata13 <- nhanes('PHTHTE_H')  # Example for phthalates and blood metals
# pfasd <- nhanes('SSPFAS_G')  # Example for phthalates and blood metals
# pfasd13 <- nhanes('SSPFAS_H')  # Example for phthalates and blood metals
vocd <- nhanes('VOCWB_G')  # Example for phthalates and blood metals
vocd13 <- nhanes('VOCWB_H')  # Example for phthalates and blood metals
hvmt <- nhanes('UAS_G')  # Example for phthalates and blood metals
hvmt13 <- nhanes('UAS_H')  # Example for phthalates and blood metals
pst <- nhanes('PSTPOL_G')  # Example for phthalates and blood metals
pst13 <- nhanes('PSTPOL_H')  # Example for phthalates and blood metals
pfc <- nhanes('PFC_G')  # Example for phthalates and blood metals
pfc13 <- nhanes('PFC_H')  # Example for phthalates and blood metals
mm <- nhanes('UHM_G')  # Example for phthalates and blood metals
mm13 <- nhanes('UHM_H')  # Example for phthalates and blood metals
cu <- nhanes('CUSEZN_G')  # Example for phthalates and blood metals
cu13 <- nhanes('CUSEZN_H')  # Example for phthalates and blood metals
ig <- nhanes('IHgEM_G')  # Example for phthalates and blood metals
ig13 <- nhanes('IHgEM_H')  # Example for phthalates and blood metals
uh <- nhanes('UHG_G')  # Example for phthalates and blood metals
uh13 <- nhanes('UHG_H')  # Example for phthalates and blood metals


merge_nhanes_data <- function( joinid, nacol,  ...) {
  datasets <- list(...)
  dfout <- Reduce(function(x, y) {
    left_join(x, y, by = joinid)  # Adjust "ID" to your common column name
  }, datasets)
  dfout = dfout[ !is.na( dfout[,nacol] ), ]
  return( dfout )
}

merged_data <- merge_nhanes_data( 'SEQN', 'CFDDS',
  merge_same_type_columns(cog_data, cog_data13),
  merge_same_type_columns(demo_data,demo_data13),
  merge_same_type_columns(depression_data,depression_data13),
  merge_same_type_columns(sleep_data,sleep_data13),
  merge_same_type_columns(bp,bp13),
  merge_same_type_columns(bm,bm13),
  merge_same_type_columns(env_data, env_data13),
  merge_same_type_columns(phdata, phdata13),
  merge_same_type_columns(vocd, vocd13),
  merge_same_type_columns(cu, cu13),
  merge_same_type_columns(ig, ig13),
  merge_same_type_columns(uh, uh13),
  mm,
#  merge_same_type_columns(pst, pst13) )
#  merge_same_type_columns(pfc, pfc13) ,
  merge_same_type_columns(hvmt, hvmt13))

table(merged_data$RIAGENDR)





idata = impute_data(merged_data)
idata = idata[ idata$BPXDI1 > 20 & idata$BPXDI2 > 20 & idata$BPXDI3 > 20 & idata$BPXDI4 > 20,  ]
fdata <- filter_na_columns(merged_data, max_na_percent = 0.5)  # Removes columns with >10% NAs
fdata = fdata[ na2f(fdata$BPXDI1 > 20 & fdata$BPXDI2 > 20 & fdata$BPXDI3 > 20 ),  ]

myform =  create_formula( fdata, 'CFDDS', c('SEQN', 'RIDAGEMN','RIDRETH3', 'CFASTAT', 'CFDCRNC', 'CFDCIT', 'CFDCST', 'CFDAPP', 'CFDAST', 'CFALANG', 'BPAARM', 'CFDCCS', 'CFDCST', 'CFDCIT', 'CFDCSR', 'CFDCIR', 'CFDCST', 'CFDCIT', 'CFDARNC',
'INDHHIN2','DMDHRBR4','INDFMIN2','PEASCST1','PEASCCT1','PEASCTM1','BPACSZ','BPXML1','URXMBP','URXMC1','URDUMMAL','SLD010H','LBXBMN','URXCNP','URXECP','URXMHH','URXMHP','URXMHNC','LBXV06','LBXV2A','LBXV3B','LBXV4C','LBXVCB','LBXVEB','LBXVCB','LBXVIPB','LBXVMC','LBXVOX','LBXVTC','LBXVTE','LBXVXY','URXUAS5','URXUAB','URXUAC','URXUMMA','URXUAS3','LBXBSE',
# getNamesFromDataframe( 'DPQ',depression_data, exclusions=c('DPQ080','DPQ030''DPQ080',)),
getNamesFromDataframe( 'DPQ',depression_data, exclusions=c('DPQ100')),
getNamesFromDataframe( 'URXUCR',fdata ),
getNamesFromDataframe( 'BMXSAD',fdata ),
getNamesFromDataframe( 'BPAEN',fdata ),
getNamesFromDataframe( 'BPXP',fdata ),
getNamesFromDataframe( 'DMD',fdata ),
getNamesFromDataframe( 'DMDHHSZ',fdata ),
getNamesFromDataframe( 'WT',fdata ),
getNamesFromDataframe( 'SDM',fdata ),
getNamesFromDataframe( 'DMQ',fdata ),
getNamesFromDataframe( 'FIA',fdata ),
getNamesFromDataframe( 'MIA',fdata ),
getNamesFromDataframe( 'AIA',fdata ),
getNamesFromDataframe( 'DMDCIT',fdata ),
getNamesFromDataframe( 'DMDYRS',fdata ),
getNamesFromDataframe( 'RIDE',fdata ),
getNamesFromDataframe( 'CFDCST',fdata ),
getNamesFromDataframe( 'SDDSRVY', fdata ),
getNamesFromDataframe( 'CFDCIT',fdata ), getNamesFromDataframe( 'LC',fdata ), getNamesFromDataframe( 'SI',fdata ) ) )
problematic_factors <- check_formula_factors(myform, df)
print(problematic_factors)
idata2=impute_data(fdata )
mdl = lm( myform, data=idata2)
summary( mdl )

```



```{r nhanessimlr}
regs <- regularizeSimlr(nh_list,fraction=0.15,sigma=rep(1.0,length(nh_list)))
regs <- regularizeSimlr(nh_list)
regs[[2]]=diag(2)
initu=initializeSimlr( nh_list, k=40, jointReduction=TRUE )
resultNH <- simlr(nh_list, iterations=100, 
        sparsenessQuantiles=rep(0.5,length(nh_list)),
        positivities=rep("positive",length(nh_list)), 
        energyType='regression', mixAlg='ica',
#        energyType='cca', mixAlg='pca',
#        scale=c("centerAndScale", "np"),
        scale=c( "robust", "centerAndScale", "np" ),
        constraint="Stiefelx1000x1000",
        initialUMatrix=initu, verbose=T)
###############
projlist=list()
mysimk=2
for ( k in 1:length(nh_list)) {
    rownames(resultNH$v[[k]])=colnames( nh_list[[k]])
    ux=data.frame(nh_list[[k]] %*% abs(resultNH$v[[k]][,1:mysimk]))
    colnames(ux)=paste0(names(nh_list)[k],1:ncol(ux))
    projlist[[length(projlist)+1]]=ux
}
simdf=simdf2=dplyr::bind_cols(projlist)
simdf2$CFDDS=nhan$CFDDS
simdf2$CFDAST=nhan$CFDAST
myform = paste0( "CFDDS ~ ",paste0( colnames(simdf), collapse='+'))
mdl = lm( myform, data=simdf2 )
print( summary( mdl ) )
#######################
interpret_simlr_vector2( resultNH$v[['exposures']], 1, n2show=5, shortnames=F )
interpret_simlr_vector2( resultNH$v[['exposures']], 2, n2show=5, shortnames=F )
# https://wwwn.cdc.gov/nchs/nhanes/search/default.aspx
# > interpret_simlr_vector2( resultNH$v[['exposures']], 1, n2show=5, shortnames=F )
#   LBXVBZ    LBXVOX    LBXVFN    LBX2DF 
# benzene, Xylene, furan, Dimethylfuran
# 1.0000000 0.3346728 0.1868323 0.0586727 
#######################
###
```
